### 2016-07-04 ansi-isolation.tex
Final set of major edits.

Fig.~\ref{fig:rel-defs} shows various relations defined over elements
in a trace. In the context of a trace $(\A,\visZ)$, an effect $\eta$
is said to belong to a transaction $T_i$ if $\eta$ belongs to the
effect set $A$ and its transaction identifier is $T_i$. The
containment relation is trivially lifted to the set of effects to
define $\underE{S \subseteq T_i}$.  Visibility and session order
relations are denoted by $\eta_1 \visar \eta_2$ and $\eta_1 \soar
\eta_2$, respectively. A transaction $T_i$ is said to be visible to an
effect $\eta$ if every effect $\eta_1$ of $T_i$ recorded by the trace
is visible to $\eta$.  $T_i$ may be visible to $\eta$ but may not be
visible to every other effect in the $\txn(\eta)$. For a transaction
$T_i$ to be considered to be visible to a transaction $T_j$ in the
context of a trace $\E$ (written $\underE{T_i \visar T_j}$), every
effect ($\eta_1$) of $T_i$ present in $\E$ must be visible to every
effect ($\eta_2$) of $T_j$ in $\E$.  Conversely, if none of the
effects of $T_i$ present in $\E$ are visible to any effect of $T_j$,
then $T_i$ is considered invisible to $T_j$ under $\E$ (written
$\underE{T_i \invisar T_j}$). Transaction $T_i$ is said to have
written to a variable $X$ under $\E$ (i.e., $\underE{T_i \wrstoar X}$)
if there exists a writes-to-$X$ effect from $T_i$ in $\E$.
\emph{Reads-from} ($\underE{T_i \rdsfmar X}$) is defined similarly.
$T_i \usesar X$ holds if $T_i$ reads or writes $X$.

>>>>

Fig.~\ref{fig:rel-defs} shows various relations defined over elements
in a trace. In the context of a trace $(\A,\visZ)$, an effect $\eta$
is said to belong to a transaction $T_i$ if $\eta$ belongs to the
effect set $A$ and its transaction identifier is $T_i$. The
containment relation is trivially lifted to the set of effects to
define $\underE{S \subseteq T_i}$.  Visibility, session order, and
happens-before relations are denoted by $\eta_1 \visar \eta_2$,
$\eta_1 \soar \eta_2$, and $\eta_1 \hboar \eta_2$, respectively. A
transaction $T_i$ is said to be visible to an effect $\eta$ if every
effect $\eta_1$ of $T_i$ recorded by the trace is visible to $\eta$.
$T_i$ may be visible to $\eta$ but may not be visible to every other
effect in the $\txn(\eta)$. For a transaction $T_i$ to be considered
to be visible to a transaction $T_j$ in the context of a trace $\E$
(written $\underE{T_i \visar T_j}$), every effect ($\eta_1$) of $T_i$
present in $\E$ must be visible to every effect ($\eta_2$) of $T_j$ in
$\E$.  Conversely, if none of the effects of $T_i$ present in $\E$ are
visible to any effect of $T_j$, then $T_i$ is considered invisible to
$T_j$ under $\E$ (written $\underE{T_i \invisar T_j}$). Transaction
$T_i$ is said to have written to a variable $X$ under $\E$ (i.e.,
$\underE{T_i \wrstoar X}$) if there exists a $\C{WR}(X)$ effect from
$T_i$ in $\E$.  \emph{Reads-from} ($\underE{T_i \rdsfmar X}$) is
defined similarly. $T_i$ \emph{uses} $X$ ($T_i \usesar X$) if it reads
or writes $X$.

### 2016-07-04 case-studies.tex

\paragraph{A video view counter.} Our first application is a Rails
implementation of a counter to count the number of views a video has
garnered on a video-sharing website like YouTube. The application
stores the video count in a MySQL database, and supports a read
operation (a database read transaction), and an increment operation (a
read-increment-write sequence wrapped in a transaction).  The
application is expected to preserve the \emph{monotonicity invariant},
namely that the view count should never appear to be decreasing.
However, the default \iso{Repeatable Read} isolation level of MySQL
leads to anomalous executions - with 32 concurrent writers performing
single increments, and one reader issuing a constant stream of reads,
we observed 12 violations of monotonicity in 10 rounds. We
subsequently instantiated our reasoning framework for an {\sc sc}
store with ANSI SQL isolation levels, and spent 5 man hours
determining appropriate isolation levels needed to enforce the
invariant. In particular, we were able to formally prove that to
preserve the invariant, writers need to execute at \iso{Serializable}
isolation, while readers can execute at \iso{Read Uncommitted}.
Repeating the experiments with this configuration led to no
violations, as expected. While executing all transactions at {\sc ser}
also prevents violations, it led to an additional 113\% increase in
the latency of writes on an average (32 concurrent writers; 5 rounds).
We repeated the experiments with the same configuration on Postgres,
and found no violations. However, Postgres's isolation
levels~\cite{postgresiso} are stronger than ANSI SQL's in non-trivial
ways. We therefore formalized Postgres' guarantees and repeated
relevant parts of the proof to discover that \iso{Repeatable Read} is
in fact sufficient for writers - an observation we confirmed through
experiments. As usual, using {\sc ser} for all transactions also
suffices, but on Postgres results in an additional 47\% increase in
write latency.

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency levels, we
re-implemented the counter as an op-based \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels (both on-demand; {\sc ec} is default), and ANSI SQL
isolation levels. Keeping the consistency level at {\sc ec}, we then
repeated our experiments with committed reads and serializable
increments - a combination which was proved to preserve monotonicity
on an {\sc sc} store, but nonetheless observed monotonicity
violations. The reason for such violations became evident when
we instantiated our proof framework for an {\sc ec} store.
Under this instantiation, it is impossible to derive the
\C{MonotonicVis} (\S\ref{sec:ansi-isolation}) property needed to show
that successive reads in a session witness monotonically increasing
state. However, assuming {\sc cc} store consistency for reads allowed
us to complete the proof, and recover the monotonicity invariant. In
fact, we were able to complete the proof even after relaxing the
serializability requirement for increments, just by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
In other words, if the reads chose the largest valued write effect,
instead of the latest write effect they witness, then counter
increments need not be serializable for reads to witness a
monotonically increasing count\footnote{The trade-off is that the view
count is now no longer accurate, although it grows monotonically.}.
We implemented this version of the counter as an op-based RDT, which
uses {\sc cc} reads and {\sc ec} increments, both un-isolated (i.e.,
\emph{Read Uncommitted}), and obtained a 52\% improvement in overall
latency compared to the version that uses serializable increments.
\vspace*{-3pt}
\paragraph{Microblog.} Our second set of experiments focused on a
Twitter-like microblogging application taken from~\cite{railsbook}.
One of the invariants that the application chooses to enforce is
referential integrity between microposts and authors. The application
enforces this invariant entirely in user space by relying on
database-backed transactions (this behavior is typical of Rails
applications~\cite{bailisferal}).  For instance, deletion of a user is
carried out in a transaction that also includes the deletion of
microposts authored by the user.  However, concurrent observers can
nonetheless witness violations of referential integrity. To measure
the likelihood of such anomalies, we first populated a Postgres
database with 1000 user accounts, each user with 50 microposts and 20
followers, chosen uniformly at random. We then constructed experiments
with 64 concurrent clients performing deletions and timeline reads of
randomly chosen users in a 1:7 ratio, and witnessed 250 violations of
referential integrity (the isolation level was left at the default
{\sc rc} level for experiments). To determine appropriate isolation
levels, we then instantiated our framework for an {\sc sc} store with
various ANSI SQL isolation levels for user deletions and timeline
reads, and obtained a proof that {\sc rc} is sufficient for deletions,
whereas reads require {\sc rr}. The proof is immediately applicable
for MySQL, which implements ANSI SQL isolation levels. Repeating our
experiments for MySQL with this configuration resulted in no
violations (as expectected), while yielding a 38\% reduction in
latency than an {\sc ser} configuration.  Since Postgres's isolation
levels are stronger than the corresponding ANSI levels,
\rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
immediately deduce that the same choice of isolation levels also work
on Postgres.

>>>>

\paragraph{A video view counter.} Our first application is a Rails
implementation of a counter to count the number of views a video has
garnered on a video-sharing website like YouTube. The application
stores the video count in a MySQL database, and supports a read
operation (a database read transaction), and an increment operation (a
read-increment-write sequence wrapped in a transaction).  The
application is expected to preserve the \emph{monotonicity invariant},
namely that the view count should never appear to be decreasing.
However, the default \iso{Repeatable Read} isolation level of MySQL
leads to anomalous executions - with 32 concurrent writers performing
single increments, and one reader issuing a constant stream of reads,
we observed 12 violations of monotonicity in 10 rounds. We
subsequently instantiated our reasoning framework for an {\sc sc}
store with ANSI SQL isolation levels, and spent 5 man hours
determining appropriate isolation levels needed to enforce the
invariant. In particular, we were able to formally prove that to
preserve the invariant, writers need to execute using
\iso{Serializable} isolation, while readers can execute using
\iso{Read Uncommitted}. Repeating the experiments with this
configuration led to no violations, as expected. While executing all
transactions at {\sc ser} also prevents violations, it led to an
additional 113\% increase in the latency of writes averaged over 5
rounds with 32 concurrent writers.  We repeated the experiments with
the same configuration on Postgres, and found no violations. However,
Postgres' isolation levels~\cite{postgresiso} are stronger than ANSI
SQL's in non-trivial ways. We therefore formalized Postgres'
guarantees and repeated relevant parts of the proof to discover that
\iso{Repeatable Read} is in fact sufficient for writers - an
observation we confirmed through experiments. As usual, using {\sc
ser} for all transactions also suffices, but on Postgres results in an
additional 47\% increase in write latency.

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing store consistency levels, we
re-implemented the counter as an op-based \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels (both on-demand; {\sc ec} is default), and ANSI SQL
isolation levels. Keeping the consistency level at {\sc ec}, we then
repeated our experiments with committed reads and serializable
increments - a combination which was proved to preserve monotonicity
on an {\sc sc} store, but nonetheless observed monotonicity
violations. The reason for such violations became evident when
we instantiated our proof framework for an {\sc ec} store.
Under this instantiation, it is impossible to derive the
\C{MonotonicVis} (\S\ref{sec:ansi-isolation}) property needed to show
that successive reads in a session witness monotonically increasing
state. However, assuming {\sc cc} store consistency for reads allowed
us to complete the proof, and recover the monotonicity invariant. In
fact, we were able to complete the proof even after relaxing the
serializability requirement for increments, just by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
In other words, if the reads chose the largest valued write effect,
instead of the latest write effect they witness, then counter
increments need not be serializable for reads to witness a
monotonically increasing count.\footnote{The trade-off is that the view
count is now no longer accurate, although it grows monotonically.}
We implemented this version of the counter as an op-based RDT, which
uses {\sc cc} reads and {\sc ec} increments, both un-isolated (i.e.,
\emph{Read Uncommitted}), and obtained a 52\% improvement in overall
latency compared to the version that uses serializable increments.
\vspace*{-3pt}

\paragraph{Microblog.} Our second set of experiments focused on a
Twitter-like microblogging application taken from~\cite{railsbook}.
An important invariant that must be preserved is the referential
integrity between microposts and authors. The application enforces
this invariant entirely in user space by relying on database-backed
transactions (this behavior is typical of Rails
applications~\cite{bailisferal}).  For instance, deletion of a user is
carried out in a transaction that also includes the deletion of
microposts authored by the user.  However, it is possible for
concurrent observers to nonetheless witness invariant violations. To
measure the likelihood of such anomalies, we first populated a
Postgres database with 1000 user accounts, each user with 50
microposts and 20 followers, chosen uniformly at random. We then
constructed experiments with 64 concurrent clients performing
deletions and timeline reads of randomly chosen users in a 1:7 ratio,
and witnessed 250 violations of referential integrity (the isolation
level was left at the default {\sc rc} level for experiments). To
determine appropriate isolation levels, we then instantiated our
framework for an {\sc sc} store with various ANSI SQL isolation levels
for user deletions and timeline reads, and obtained a proof that {\sc
rc} is sufficient for deletions, whereas reads require {\sc rr}. The
proof is immediately applicable for MySQL, which implements ANSI SQL
isolation levels. Repeating our experiments for MySQL with this
configuration resulted in no violations (as expected), while yielding
a 38\% reduction in latency than an {\sc ser} configuration.  Since
Postgres's isolation levels are stronger than the corresponding ANSI
levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
immediately deduce that the same choice of isolation levels also work
on Postgres.

### 2016-07-04 motivation.tex

As an alternative, consider the execution of this transaction under
\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} that read uncommitted
data. In the current example, {\sc rc} isolation admits the two
executions shown in Fig.~\ref{fig:rc-ex} on a strongly consistent
({\sc sc}) store, such as a relational database.

@@

The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it. In the first execution, transaction \C{Wd1} reads
the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
before it commits transaction \C{Wd2} executes and commits, writing
the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
witnessing the uncommitted writes of transaction \C{Wd1}.
Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
in an incorrect balance of \C{k-a1}. The second execution describes a
similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
Clearly, {\sc rc} is an excessively weak isolation level for this
program because it loses the updates of one transaction, resulting in
the violation of the post-condition.  A stronger isolation level that
prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} however does not serialize
transactions in the absence of write-write conflicts.} Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based concurrency
control, it is also more efficient, making it appropriate for both
\C{Wd} transactions.

@@

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}, a compositional proof technique that
allows us to reason about the behavior of individual threads by
abstracting away interferences induced by other threads (collectively
called \emph{the environment}) into a \emph{rely} relation. In an
ordinary concurrent program, every environment step is a valid
interference in the current thread. However, in the presence of
transactions executing under various levels of weak isolation,
determining what constitutes an interference is a non-trivial problem.
For example, a \iso{Serializable} transaction admits no interference,
whereas within a transaction executing under \iso{Read Uncommitted}
isolation all interferences are valid. Between these two extremes are
various levels of isolation that admit some interferences while
prohibiting others. For example, \iso{Read Committed} isolation admits
interference of a committed transaction, but not that of an
uncommitted transaction. \iso{Snapshot Isolation} admits interference
from committed non-conflicting transactions, and so on.

@@

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the definition of the
\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
what actions become visible to a transaction, or its constituents, and
when. To do so, we axiomatize the $\hbZ$ relation to capture the
interference characteristics of different isolation levels as their
specifications.  Two such specifications are shown\footnote{Note that
specifications presented here are solely for illustration. The actual
specifications described in \S\ref{sec:ansi-isolation} are more
nuanced.} in Fig.~\ref{fig:interference-ex}.  \iso{Read Committed}
specification ($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to
happen before an operation $\eta_2$ of $T_2$ only if every operation
in $T_1$ (including its commit) happens before $\eta_2$ (we let $T_1
\hboar \eta_2$ denote this). The specification does not require $T_1$
to execute and commit before all operations of $T_2$, thus allowing it
to interfere in $T_2$. The \iso{Snapshot Isolation} specification
($\psi_{SI}$) requires two transactions that write to the same
variable to be related by happens-before.  This effectively prohibits
interference due to actions in $T_1$ being interleaved in $T_2$ (or,
vise versa).

@@

The specification of an isolation level encodes its interference
characteristics as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or invalid, allowing the programmer to only
focus on the former. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{C} in \C{Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):

\begin{smathpar}
\begin{array}{l}
  \neg\committed(\C{Wd1}) \conj (\neg\committed(\C{Wd2}) \Rightarrow
  \C{C = k-a1}) 
                \\
      \C{Wd1} \wrstoar \C{C} \conj (\committed(\C{Wd2})
                \Rightarrow \C{C = k-a1-a2})
\end{array}
\end{smathpar}

>>>>

As an alternative, consider the execution of this transaction under
a \emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}.\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} that read uncommitted
data. In the current example, {\sc rc} isolation admits the two
executions shown in Fig.~\ref{fig:rc-ex} on a strongly consistent
({\sc sc}) store, such as a relational database.

@@

The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it. In the first execution, transaction \C{Wd1} reads
the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
before it commits, transaction \C{Wd2} executes and commits, writing
the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
witnessing the uncommitted writes of transaction \C{Wd1}.
Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
(the so-called \emph{lost update anomaly}~\cite{berenson}), resulting
in an incorrect balance of \C{k-a1}. The second execution describes a
similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
Clearly, {\sc rc} is an excessively weak isolation level for this
program because it loses the updates of one transaction, resulting in
the violation of the post-condition.  A stronger isolation level that
prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} however does not serialize
transactions in the absence of write-write conflicts.} Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based concurrency
control, it is also more efficient, making it appropriate for both
\C{Wd} transactions.

@@

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}, a compositional proof technique that
allows us to reason about the behavior of individual threads by
abstracting away interferences induced by other threads (collectively
called \emph{the environment}) into a \emph{rely} relation. In an
ordinary concurrent program, every environment step is a valid
interference in the current thread. However, in the presence of
transactions executing under various levels of weak isolation,
determining what constitutes an interference is a non-trivial problem.
For example, a \iso{Serializable} transaction admits no interference,
whereas within a transaction executing under \iso{Read Uncommitted}
isolation, all interferences are valid. Between these two extremes are
various levels of isolation that admit some interferences while
prohibiting others. For example, \iso{Read Committed} isolation admits
interference of a committed transaction, but not those from
uncommitted transactions. \iso{Snapshot Isolation} admits interference
from committed non-conflicting transactions, and so on.

@@

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the definition of the
\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
what actions become visible to a transaction, or its constituents, and
when. To do so, we axiomatize the $\hbZ$ relation to capture the
interference characteristics of different isolation levels as their
specifications. Two such specifications are shown\footnote{Note that
specifications presented here are solely for illustration. The actual
specifications described in \S\ref{sec:ansi-isolation} are more
nuanced.} in Fig.~\ref{fig:interference-ex}. The \iso{Read Committed}
specification ($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to
happen-before an operation $\eta_2$ of $T_2$ only if every operation
in $T_1$ (including its commit) happens-before $\eta_2$ (we let $T_1
\hboar \eta_2$ denote this). The specification does not require $T_1$
to execute and commit before all operations of $T_2$, thus allowing it
to interfere in $T_2$. The \iso{Snapshot Isolation} specification
($\psi_{SI}$) requires two transactions that write to the same
variable to be related by happens-before.  This effectively prohibits
interference due to actions in $T_1$ being interleaved in $T_2$ (or,
vise versa).

@@

The specification of an isolation level encodes its interference
characteristics as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or not, allowing the programmer to only
focus on the former. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{C} in \C{Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):

\begin{smathpar}
\begin{array}{c}
 \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C}  \conj 
    (\neg\committed(\C{Wd2}) \Rightarrow \C{C = k-a1}) \\
    \conj (\committed(\C{Wd2})
                \Rightarrow \C{C = k-a1-a2})
\end{array}
\end{smathpar}

### 2016-07-04 reasoning.tex

\noindent The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
shared variable reads inside a transaction $T_i$. It requires $\I$ to
be stable, and pre- and post- conditions to be stable relative to
$\I$.  The quantified premise effectively requires a proof that if the
abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
an execution $\E$ that satisfies the pre-condition $P$, then the
resultant execution $\E'$ satisfies the post-condition $Q$, and that
the guarantee $G$ faithfully captures the transition from $\E$ to
$\E'$. It is informative to compare this premise with the premise of
the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
premises also appear in \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
rules, which define rely-guarantee judgments for assignments and
transactions, respectively. \rulelabel{RG-Var} however also requires
the return value ($\interp{S}(X)$) of the read to satisfy the
assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the RG
judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
corresponding judgments for the constituent expressions $e_1$ and
$e_2$. The quantified premise requires any value resulting from
evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
$e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
$C_2$, respectively. The rules for sequential and parallel composition
of commands are essentially the same as their counterparts in a
standard rely-guarantee formulation and hence elided.

@@

The \rulelabel{RG-Conseq} rule define ways to strengthen or
weaken relations and assertions associated with the RG judgment of
transaction-bound expressions. Similar rules exist for
transaction-bound and top-level commands, but are not
shown.\footnote{The supplementary materials provide the complete set
  of rules.} As is the case with a standard rely-guarantee
formulation,the rules allow the pre-condition $P$ and the rely relation
$R$ to be strengthened, and the post-condition $Q$ (also, $C$ in the
case of expressions) and the guarantee relation $G$ to be weakened.
The most notable aspect of the \rulelabel{RG-Conseq} rules is that
they allow the trace invariant $\I$ to be strengthened. Considering
that $\I$ captures isolation properties, this means that a program
proven correct under weaker isolation levels is also correct under
stronger ones.  Parametricity over the trace invariant $\I$, combined
with the ability to strengthen $\I$ as needed, allows our proof system
to support a highly flexible proof strategy to prove programs correct
over various isolation variants. For example, programmers can
\emph{define} isolation guarantees of their choice (by defining $\I$
appropriately) and then prove programs correct assuming the guarantees
hold.  The soundness of strengthening $\I$ ensures that a program can
be executed safely on any data store that offers isolation guarantees
at least as strong as those assumed.

@@

The interleaved step relation for transaction bound expressions
($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
reflexive transitive closure of the interleaved step relation.
\end{definition}

>>>>

\noindent The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
shared variable reads inside a transaction $T_i$. It requires $\I$ to
be stable, and pre- and post- conditions to be stable relative to
$\I$.  The quantified premise effectively requires a proof that if the
abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
an execution $\E$ that satisfies the pre-condition $P$, then the
resultant execution $\E'$ satisfies the post-condition $Q$, and that
the guarantee $G$ faithfully captures the transition from $\E$ to
$\E'$. It is informative to compare this premise with the premise of
the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
premises also appear in the \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
rules, which define rely-guarantee judgments for assignments and
transactions, respectively. \rulelabel{RG-Var} however also requires
the return value ($\interp{S}(X)$) of the read to satisfy the
assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the RG
judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
corresponding judgments for the constituent expressions $e_1$ and
$e_2$. The quantified premise requires any value resulting from
evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
$e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
$C_2$, respectively. The rules for sequential and parallel composition
of commands are essentially the same as their counterparts in a
standard rely-guarantee formulation and hence elided.

@@

The \rulelabel{RG-Conseq} rule define ways to strengthen or
weaken relations and assertions associated with the RG judgment of
transaction-bound expressions. Similar rules exist for
transaction-bound and top-level commands, but are not
shown.\footnote{The supplementary materials provide the complete set
  of rules.} As is the case with a standard rely-guarantee
formulation,the rules allow the pre-condition $P$ and the rely relation
$R$ to be strengthened, and the post-condition $Q$ (also, $C$ in the
case of expressions) and the guarantee relation $G$ to be weakened.
The most notable aspect of the \rulelabel{RG-Conseq} rules is that
they allow the trace invariant $\I$ to be strengthened. Considering
that $\I$ captures isolation properties, this means that a program
proven correct under weaker isolation levels is also correct under
stronger ones.  Parametricity over the trace invariant $\I$, combined
with the ability to strengthen $\I$ as needed, allows our proof system
to support a highly flexible proof strategy to prove programs correct
over various isolation variants. For example, programmers can
\emph{define} isolation guarantees of their choice (by defining $\I$
appropriately) and then prove programs correct assuming the guarantees
hold.  The soundness of strengthening $\I$ ensures that a program can
be safely executed on any system that offers isolation guarantees
at least as strong as those assumed.

@@

\noindent The interleaved step relation for transaction bound expressions
($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
reflexive transitive closure of the interleaved step relation.
\end{definition}

### 2016-07-04 related-work.tex

\paragraph{Reasoning under weak consistency} There have been several
recent proposals to reason about programs executing under weak
consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
  redblueosdi, ecinec}. All of them assume a system model that offers
a choice between a \emph{coordination-free} weak consistency level
(\emph{e.g.}, eventual consistency~\cite{redblueosdi, redblueatc,
  ecinec, alvarocalm, bailisvldb}) or causal
consistency~\cite{LBC16,gotsmanpopl16}). The key technical issue in
these efforts involves proving that atomic and fully isolated
operations preserve application invariants when executed on these
consistency levels.  In contrast, we admit weakly isolated
transactions, and our system model accepts \emph{specifications} of
consistency and isolation levels drawn from an expressive logic.
Gotsman \emph{et al.}~\cite{gotsmanpopl16} adapts \iso{Parallel
  Snapshot Isolation} to the aforementioned setting by interpreting it
as a consistency level that serializes writes to objects; a dedicated
proof rule is developed to help prove prove program invariants hold
under this model. By parameterizing our proof system over a gamut of
weak isolation specifications, we avoid the need to define a separate
proof rule for each new isolation level we may encounter.

>>>>

\paragraph{Reasoning under weak consistency} There have been several
recent proposals to reason about programs executing under weak
consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
redblueosdi, ecinec}. All of them assume a system model that offers a
choice between a \emph{coordination-free} weak consistency level
(\emph{e.g.}, eventual consistency~\cite{redblueosdi, redblueatc,
ecinec, alvarocalm, bailisvldb}) or causal
consistency~\cite{lbc16,gotsmanpopl16}). The key technical issue in
these efforts involves proving that atomic and fully isolated
operations preserve application invariants when executed on these
consistency levels.  In contrast, we admit weakly isolated
transactions, and our system model accepts \emph{specifications} of
consistency and isolation levels drawn from an expressive logic.
Gotsman \emph{et al.}~\cite{gotsmanpopl16} adapts \iso{Parallel
Snapshot Isolation} to the aforementioned setting by interpreting it
as a consistency level that serializes writes to objects; a dedicated
proof rule is developed to help prove prove program invariants hold
under this model. By parameterizing our proof system over a gamut of
weak isolation specifications, we avoid the need to define a separate
proof rule for each new isolation level we may encounter.
